---
title: "Hawaii Mycobiome"
author: "Laura Tipton"
date: "October 25, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(vegan)
require(hillR)
require(vioplot)
require(car)
require(RColorBrewer)
require(sads)
#load data

```

In the data loaded, the following data frames are loaded:
  esvs contains the ESV by sample table
  otus contains the OTU by sample table, with samples in the same order as esvs
  mapping contains the 'metadata' for the samples
  ESVbasedOTUs which contains OTUs in rows and ESVs in columns with 1's to indicate which ESVs are in each OTU

Additionally, the following vectors have already been isolated from the mapping data frame:
  ecocol is the samples coded (as colors) by their ecosystem
  islcol is the samples coded (as colors) by their island
  runcol is the samples coded (as colors) by their survey
See legends for color mappings

Finally, one personal function is imported
  gt0 counts the number of instances greater than 0 in a vector
  
Separate large OTUs and create bar plots for their islands of origin
```{r large OTUs}
#denovo9552 (contains most ESVs)
c9552 <- colnames(ESVbasedOTUs)[which(ESVbasedOTUs[9553,]==1)]
e9552 <- as.numeric(substr(c9552, 13, 20))
denovo9552.temp <- esvs[,e9552]
denovo9552 <- denovo9552.temp[-which(rowSums(denovo9552.temp)<10),]
d9552ecocol <- ecocol[-which(rowSums(denovo9552.temp)<10)]
d9552islcol <- islcol[-which(rowSums(denovo9552.temp)<10)]
d9552runcol <- runcol[-which(rowSums(denovo9552.temp)<10)]

#denovo7896 ()
c7896 <- colnames(ESVbasedOTUs)[which(ESVbasedOTUs[7897,]==1)]
e7896 <- as.numeric(substr(c7896, 13, 20))
denovo7896.temp <- esvs[,e7896]
denovo7896 <- denovo7896.temp[-which(rowSums(denovo7896.temp)<10),]
d7896ecocol <- ecocol[-which(rowSums(denovo7896.temp)<10)]
d7896islcol <- islcol[-which(rowSums(denovo7896.temp)<10)]
d7896runcol <- runcol[-which(rowSums(denovo7896.temp)<10)]

#denovo10518 ()
c10518 <- colnames(ESVbasedOTUs)[which(ESVbasedOTUs[10519,]==1)]
e10518 <- as.numeric(substr(c10518, 13, 20))
denovo10518.temp <- esvs[,e10518]
denovo10518 <- denovo10518.temp[-which(rowSums(denovo10518.temp)<10),]
d10518ecocol <- ecocol[-which(rowSums(denovo10518.temp)<10)]
d10518islcol <- islcol[-which(rowSums(denovo10518.temp)<10)]
```

```{r figure 1}
#figure 1A denovo9552
d9552byisleco <- aggregate(denovo9552, list(d9552islcol, d9552ecocol), sum)
rownames(d9552byisleco) <- paste(d9552byisleco[,1], d9552byisleco[,2])
d9552byisleco <- d9552byisleco[-c(5),-c(1:2)] #remove unassigned islands and/or ecosystems (row 5)
#convert to relative abundance
d9552islecosums <- apply(d9552byisleco, 2, sum)
d9552islecorel <- matrix(0, nrow=nrow(d9552byisleco), ncol=ncol(d9552byisleco))
for(i in 1:ncol(d9552byisleco)){
  d9552islecorel[,i] <- d9552byisleco[,i]/d9552islecosums[i]
}
colnames(d9552islecorel) <- colnames(d9552byisleco)
#plot
barplot(cbind(rowSums(d9552byisleco[order(rownames(d9552byisleco)),])/sum(d9552byisleco),
              d9552islecorel[order(rownames(d9552byisleco)),]), 
        col=c("darkgreen", "goldenrod", "olivedrab", "turquoise", "olivedrab", "chocolate", "goldenrod", "olivedrab"),
        names.arg=c("OTU9552",substr(c9552,10,20)), las=2, lwd=2, 
        space=c(0.2, 0.7, rep(0.2,22)), density=c(10, 10, 10, 10, 17, 20, 20, 20), 
        angle=c(45, 45, 45, 45, 90, 315, 315, 315), border=NA)

#figure 1B denovo10518
d10518byisleco <- aggregate(denovo10518, list(d10518islcol, d10518ecocol), sum)
rownames(d10518byisleco) <- paste(d10518byisleco[,1], d10518byisleco[,2])
d10518byisleco <- d10518byisleco[-c(1,2,7),-c(1:2)] #remove unknown island and/or ecosystem, in rows 1, 2, and 7
#convert to relative abundance
d10518islecosums <- apply(d10518byisleco, 2, sum)
d10518islecorel <- matrix(0, nrow=nrow(d10518byisleco), ncol=ncol(d10518byisleco))
for(i in 1:ncol(d10518byisleco)){
  d10518islecorel[,i] <- d10518byisleco[,i]/d10518islecosums[i]
}
colnames(d10518islecorel) <- colnames(d10518byisleco)
#plot
barplot(cbind(rowSums(d10518byisleco[order(rownames(d10518byisleco)),])/sum(d10518byisleco),
              d10518islecorel[order(rownames(d10518byisleco)),]),
        col=c("darkgreen", "goldenrod", "olivedrab", "olivedrab", "turquoise", "chocolate", "goldenrod", "olivedrab"),
        names.arg=c("OTU10518",substr(c10518,10,20)), las=2, lwd=2, 
        space=c(0.2, 0.7, rep(0.2,11)), density=c(10, 10, 10, 17, 17, 20, 20, 20),
        xlim=c(0,23), angle=c(45, 45, 45, 90, 90, 315, 315, 315), border=NA)
#add legend to this one because of room
legend("right", fill=c("darkgreen", "goldenrod", "turquoise", "olivedrab", "chocolate", "black", "black", "black"),
       angle=c(rep(NULL, 5), 90, 315, 45), density=c(rep(1000, 5), 10, 17, 20), bty="n",
       cex=2,legend=c("Endophyte", "Aerobiota", "Marine", "Epiphyte", "Soil", "Oahu", "Maui Nui", "Hawaii"))

#figure 1C denovo7896
d7896byisleco <- aggregate(denovo7896, list(d7896islcol, d7896ecocol), sum)
rownames(d7896byisleco) <- paste(d7896byisleco[,1], d7896byisleco[,2])
d7896byisleco <- d7896byisleco[-c(1,2,8,12),-c(1:2)] #remove rows with unknown island or ecosystem, rows 1, 2, 8, and 12
#convert to relative abundance
d7896islecosums <- apply(d7896byisleco, 2, sum)
d7896islecorel <- matrix(0, nrow=nrow(d7896byisleco), ncol=ncol(d7896byisleco))
for(i in 1:ncol(d7896byisleco)){
  d7896islecorel[,i] <- d7896byisleco[,i]/d7896islecosums[i]
}
colnames(d7896islecorel) <- colnames(d7896byisleco)
#plot
barplot(cbind(rowSums(d7896byisleco[order(rownames(d7896byisleco)),])/sum(d7896byisleco), d7896islecorel[order(rownames(d7896byisleco)),]),
        col=c("chocolate", "darkgreen", "goldenrod", "olivedrab", "turquoise", "olivedrab", "turquoise", "chocolate", "goldenrod", "olivedrab"),
        names.arg=c("OTU7896",substr(c7896,10,20)), las=2, lwd=2, 
        space=c(0.2, 0.7, rep(0.2,13)), density=c(10, 10, 10, 10, 10, 17, 17, 20, 20, 20),
        xlim=c(0,23), angle=c(45, 45, 45, 45, 45, 90, 90, 315, 315, 315), border=NA)

```

Calculate diversity metrics. For beta diversity, samples with a single OTU are removed, as they end up being outliers in the MDS plots

```{r diversity metrics}

#alpha diversity
otush0 <- hill_taxa(otus, 0, MARGIN=1)
otush1 <- hill_taxa(otus, 1, MARGIN=1)
otush2 <- hill_taxa(otus, 2, MARGIN=1)
esvsh0 <- hill_taxa(esvs, 0, MARGIN=1)
esvsh1 <- hill_taxa(esvs, 1, MARGIN=1)
esvsh2 <- hill_taxa(esvs, 2, MARGIN=1)

#beta diversity
notus <- apply(otus, 1, gt0)

otujac <- metaMDS(otus[-c(which(notus==1),290),], distance="jaccard", binary=TRUE, trymax=200)
esvjac <- metaMDS(esvs[-c(which(notus==1),290),], distance="jaccard", binary=TRUE, trymax=200)

```


```{r fig 2}
#violin plots

vioplot(otush0~recode(runcol, '"#8DD3C7" = "BIG";
"#FFFFB3" = "FEF2"; "#BEBADA" = "FEF3"; "#FB8072" = "MesoAlgae"; "#80B1D3" = "MLO";
"#FDB462" = "Nguen"; "#B3DE69" = "PeayVitousek"; "#FCCDE5" = "Swabs";
"#D9D9D9" = "Lau"; "#BC80BD" = "Kipuka"; "darkgrey" = "FEF3"'),
col=c(brewer.pal(10,"Set3")[c(1,2,3,10,9,4,5,6,7,8)]),
at=seq(1,20,2), wex=c(382, 276, 252, 80, 335, 23, 276, 5, 156, 190)/382,
ylim=c(0, 720), ylab="Observed Richness", xlab="Survey")

vioplot(esvsh0~recode(runcol, '"#8DD3C7" = "BIG";
"#FFFFB3" = "FEF2"; "#BEBADA" = "FEF3"; "#FB8072" = "MesoAlgae"; "#80B1D3" = "MLO";
"#FDB462" = "Nguen"; "#B3DE69" = "PeayVitousek"; "#FCCDE5" = "Swabs";
"#D9D9D9" = "Lau"; "#BC80BD" = "Kipuka"; "darkgrey" = "FEF3"'),
border=c(brewer.pal(10,"Set3")[c(1,2,3,10,9,4,5,6,7,8)]),
rectCol=c(brewer.pal(10,"Set3")[c(1,2,3,10,9,4,5,6,7,8)]),
lineCol=c(brewer.pal(10,"Set3")[c(1,2,3,10,9,4,5,6,7,8)]),
at=seq(1.75,20.75,2), wex=c(382, 276, 252, 80, 335, 23, 276, 5, 156, 190)/382,
add=TRUE, col=NA, ylab="Observed Richness", xlab="Survey", colMed="black")

```

```{r fig 3}
#procrustes plot

proc <- procrustes(otusjac$points, esvsjac$points)
plot(proc)
```

```{r fig 4}
#octave plot

octavotus <- octav(colSums(otus))
octavesvs <- octav(colSums(esvs))

plot(otctavesvs, prop=TRUE, border=NA, col=NA)
lines(octavsesvs, prop=TRUE, mid=FALSE, col="red", pch=19)
lines(octavsotus, prop=TRUE, mid=FALSE, col="blue", pch=19)

ks.test(octavotus$Freq, octavesvs$Freq)